openarchaeoCollaboration

### 2023-06-14 analysis/_03-data_methodology.qmd
Merge code chunks into section qmds
Leaving the network analysis unevaluated for now.

@@

{r plot-repo-hosts}

**NOTE:** Indicate the numerical value at the end of each bar in the plot-repo-hosts figure.

{r table-no-repo}

>>>>

```{r data-oarch}
read_csv("https://raw.githubusercontent.com/zackbatist/open-archaeo/master/open-archaeo.csv") %>%
  mutate(category = na_if(category, "?")) ->
  oarch

# Extract clean repos from GitHub URLs
oarch %>%
  mutate(
    gh_repo = str_match(oarch$github, "^https://github.com/([\\w-\\.]+/[\\w-\\.]+)/?.*?$")[,2]
  ) ->
  oarch

# Amalgamate categories into fewer rough categories
# TODO: redo
oarch <- mutate(oarch, rough_category = recode(category,
                                 "R" = "Packages (R, Python, etc.)",
                                 "Web apps" = "Standalone apps",
                                 "Datasets" = "Analyses & datasets",
                                 "Tutorials" = "Lists & tutorials",
                                 "Python" = "Packages (R, Python, etc.)",
                                 "Desktop apps" = "Standalone apps",
                                 "QGIS" = "Plugins",
                                 "Lists" = "Lists & tutorials",
                                 "Mobile apps" = "Standalone apps",
                                 "Analyses" = "Analyses & datasets",
                                 "Spreadsheets" = "Forms & spreadsheets",
                                 "Command line tools" = "Standalone apps",
                                 "Plugins" = "Plugins",
                                 "ArcGis" = "Plugins",
                                 "Filemaker" = "Plugins",
                                 "ODK" = "Forms & spreadsheets",
                                 "Lisp" = "Packages (R, Python, etc.)",
                                 .default = "Other",
                                 .missing = "Other"
                                 ))
```

@@

```{r data-github}
# If cached data at `analysis/data/derived_data/oarch.RData` is present, this
# chunk will prefer to load it rather than make requests to the GitHub API.
# Delete this file to fetch fresh data from the GitHub or modify 
# `cached_gh_data` to read a different cached version.
#
# Note that fetching data from the GitHub API takes some time and is subject to
# a rate limit:
# <https://docs.github.com/en/rest/overview/resources-in-the-rest-api#rate-limiting>
#
# You will need to authenticate authenticate with GitHub credentials to get a
# rate limit sufficient to obtain the required data, see 
# <https://gh.r-lib.org/articles/managing-personal-access-tokens.html>
# for options on how to do this.
cached_gh_data <- here("analysis/data/derived_data/oarch.RData")
if (!file_exists(cached_gh_data)) {
  oarch <- mutate(oarch,
                  gh_langs = map(oarch$gh_repo, gh_lang),
                  gh_contribs = map(oarch$gh_repo, gh_contrib),
                  gh_issues = map(oarch$gh_repo, gh_issue),
                  gh_comments = map(oarch$gh_repo, gh_comment),
                  gh_commits = map(oarch$gh_repo, gh_commit)
  )
  saveRDS(oarch, cached_gh_data)
}
oarch <- readRDS(cached_gh_data)
```

```{r plot-repo-hosts}
# TODO: turn into table or just text
# TODO: Indicate the numerical value at the end of each bar in the plot-repo-hosts figure.
oarch <- mutate(oarch, none = ifelse(is.na(github) &
                                       is.na(gitlab) &
                                       is.na(bitbucket) &
                                       is.na(launchpad) &
                                       is.na(codeberg),
                                     "", NA))

oarch %>%
  pivot_longer(c(github, gitlab, bitbucket, launchpad, codeberg, none),
               names_to = "repo_host", values_to = "repo_url") %>%
  drop_na(repo_url) %>%
  group_by(repo_host, author1_name) %>%
  summarise(n = n(), .groups = "drop_last") %>%
  mutate(total = sum(n)) %>%
  ungroup() %>%
  mutate(repo_host = fct_reorder(repo_host, total),
         repo_host = fct_relevel(repo_host, "none")) %>%
  ggplot(aes(repo_host, n, group = author1_name)) +
  geom_col(colour = "#333333") +
  coord_flip() +
  labs(title = "Most popular repository hosts",
       subtitle = "By number of projects, split by primary author",
       caption = "Data: open-archaeo.info",
       x = NULL, y = NULL)
```

```{r table-no-repo}
# TODO: prettify with gt()
filter(oarch, none == "") %>%
  count(category)
```

### 2023-06-14 analysis/_04-what_is_open_archaeology.qmd
Merge code chunks into section qmds
Leaving the network analysis unevaluated for now.

{r plot-categories}

@@

{r plot-platforms}

@@

{r plot-tags-cats}

>>>>

```{r plot-categories}
oarch <- mutate(oarch,
                category = recode(category,
                                  "Specifications, protocols and schemas" = "Specifications etc.",
                                  .default = category))
oarch %>%
  drop_na(category) %>%
  group_by(category) %>%
  summarise(n = n()) %>%
  ggplot(aes(fct_reorder(category, n), n)) +
  geom_col() +
  coord_flip() +
  labs(
    title = "Most popular categories",
    subtitle = "By number of projects",
    caption = "Data: open-archaeo.info",
    x = NULL, y = NULL
  ) +
  theme_oarch()
```

@@

```{r plot-platforms}
oarch %>%
  drop_na(platform) %>%
  group_by(platform, author1_name) %>%
  summarise(n = n(), .groups = "drop_last") %>%
  mutate(total = sum(n)) %>%
  ggplot(aes(fct_reorder(platform, total), n, group = author1_name)) +
  geom_col(colour = "#333333") +
  coord_flip() +
  labs(
    title = "Most popular platforms",
    subtitle = "By number of projects, split by primary author",
    caption = "Data: open-archaeo.info",
    x = NULL, y = NULL
  ) +
  theme_oarch()
```

@@

```{r plot-tags-cats}
detail_tags <- c("Instrumental Neutron activation analysis",
                 "Harris Matrix",
                 "aDNA Simulators",
                 "Damage profilers",
                 "Metagenomic classifers",
                 "aDNA simulators",
                 "aDNA damage-aware genotypers",
                 "Low-coverage contamination estimation",
                 "Low-coverage adapted population-genetics tools",
                 "aDNA preprocessing",
                 "aDNA aligner",
                 "aDNA assembler",
                 "aDNA toolkit")

oarch %>%
  pivot_longer(c(tag1, tag2, tag3, tag4, tag5), values_to = "tag",
               values_drop_na = TRUE) %>%
  drop_na(tag) %>%
  filter(!tag %in% detail_tags) %>%
  mutate(category = recode(category,
                                  "Specifications etc." = "Documents",
                                  "Products" = "Documents",
                                  "Guides" = "Documents",
                                  "Lists and datasets" = "Documents",
                                  .default = category)) %>%
  group_by(tag, category) %>%
  summarise(n = n(), .groups = "drop_last") %>%
  mutate(total = sum(n)) %>%
  ggplot(aes(fct_reorder(tag, total), n, fill = category)) +
  # facet_wrap(vars(category)) +
  geom_col() +
  scale_fill_bright() +
  coord_flip() +
  labs(
    title = "Tags by category",
    caption = "Data: open-archaeo.info",
    x = NULL, y = "Number of projects",
    fill = NULL
  ) +
  theme_oarch() +
  theme(
    axis.text.y = element_text(size = 7)
  )
```

### 2023-06-14 analysis/_05-history_of_open_archaeology.qmd
Merge code chunks into section qmds
Leaving the network analysis unevaluated for now.

{r plot-gh-cum}

>>>>

```{r plot-gh-cum}
oarch %>% 
  mutate(
    lumped_category = recode(category,
                             "Specifications etc." = "Documents",
                             "Products" = "Documents",
                             "Guides" = "Documents",
                             "Lists and datasets" = "Documents",
                             .default = category),
    lumped_category = factor(lumped_category, c("Scripts",
                                                "Documents",
                                                "Packages and libraries",
                                                "Standalone software")) ) ->
  oarch_lumped

p_gh_commits_cum <- oarch_lumped %>%
  unnest(gh_commits) %>%
  drop_na(sha) %>%
  mutate(date = floor_date(datetime, "week")) %>%
  group_by(lumped_category, date) %>%
  summarise(n_commits = n(), .groups = "drop_last") %>%
  mutate(cum_commits = cumsum(n_commits)) %>%
  ggplot(aes(date, cum_commits,
             fill = lumped_category,
             colour = lumped_category,
             label = lumped_category)) +
  geom_area() +
  annotate("segment", 
           x = ymd("2008-04-01", tz = "UTC"), 
           xend = ymd("2008-04-01", tz = "UTC"),
           y = 16000, yend = 4000,
           arrow = arrow(type = "closed", length = unit(0.02, "npc"))) +
  annotate("text", 
           x = ymd("2008-04-01", tz = "UTC"), y = 19000,
           label = "GitHub\nlaunched",
           vjust = 0) +
  scale_x_datetime(date_breaks = "2 years", date_labels = "%Y") +
  scale_y_continuous(labels = scales::comma) +
  scale_colour_bright(guide = guide_none()) +
  scale_fill_bright() +
  labs(
    y = "cumulative commits",
    x = NULL, fill = NULL
  ) +
  theme_oarch() +
  theme(legend.position = "top")

# TODO: extend to end
p_gh_repos_cum <- oarch_lumped %>%
  unnest(gh_commits) %>%
  drop_na(sha) %>%
  group_by(repo) %>% 
  filter(datetime == min(datetime)) %>% # Only first commits
  mutate(date = floor_date(datetime, "week")) %>%
  group_by(lumped_category, date) %>%
  summarise(n_commits = n(), .groups = "drop_last") %>%
  mutate(cat_total = sum(n_commits)) %>%
  mutate(cum_commits = cumsum(n_commits)) %>%
  ggplot(aes(date, cum_commits,
             fill = lumped_category,
             colour = lumped_category)) +
  geom_area() +
  annotate("segment",
           x = ymd("2008-04-01", tz = "UTC"),
           xend = ymd("2008-04-01", tz = "UTC"),
           y = 40, yend = 10,
           arrow = arrow(type = "closed", length = unit(0.02, "npc"))) +
  annotate("text",
           x = ymd("2008-04-01", tz = "UTC"), y = 50,
           label = "GitHub\nlaunched",
           vjust = 0) +
  scale_x_datetime(date_breaks = "2 years", date_labels = "%Y") +
  scale_y_continuous(labels = scales::comma) +
  scale_colour_bright(guide = guide_none()) +
  scale_fill_bright(guide = guide_none()) +
  labs(
    y = "cumulative repositories",
    x = NULL, fill = NULL
  ) +
  theme_oarch()

p_gh_commits_cum / p_gh_repos_cum
```

### 2023-06-14 analysis/_06-project_activity.qmd
Merge code chunks into section qmds
Leaving the network analysis unevaluated for now.

{r plot-lifespan}
{r lifespan-stats}

@@

**NOTE:** Combine these figures using patchwork.
{r plot-lifespan-hist}
{r plot-lifespan-vs-rate}

@@

{r plot-gh_contributors-histogram}

@@

{r plot-contrib-dist}

@@

**NOTE:** Combine these figures using patchwork.
{r plot-issues-hist}
{r plot-comments-hist}

>>>>

```{r plot-lifespan}
oarch %>%
  unnest(gh_commits) %>%
  drop_na(datetime) %>%
  group_by(item_name) %>%
  add_tally() %>%
  filter(n > 1) %>%
  mutate(first_commit = min(datetime),
         mid_commit = median(datetime),
         last_commit = max(datetime)) %>%
  ggplot(aes(x = fct_reorder(item_name, last_commit, .desc = FALSE),
             y = datetime,
             ymin = first_commit,
             ymax = last_commit)) +
  geom_linerange(size = 0.05) +
  geom_point(size = 0.1) +
  scale_y_datetime(limits = c(ymd("2005-01-01", tz = "UTC"),
                              ymd("2021-05-31", tz = "UTC")),
                   date_breaks = "2 years", date_labels = "%Y") +
  # coord_flip() +
  labs(
    title = "Project lifespan",
    subtitle = glue("Individual commits in GitHub repositories (with more than 1 commit)"),
    x = NULL, y = NULL,
    caption = "Data: open-archaeo.info and GitHub API"
  ) +
  theme_oarch() +
  theme(
    axis.line.x = element_blank(),
    axis.ticks.x = element_blank(),
    axis.text.x = element_blank(),
    panel.grid.minor.x = element_blank(),
    panel.grid.major.x = element_blank()
  )
```

```{r lifespan-stats}
oarch %>% 
  unnest(gh_commits) %>% 
  drop_na(datetime) %>% 
  group_by(item_name) %>%
  mutate(n_contribs = map_int(gh_contribs, vctrs::vec_size)) %>% 
  summarise(
    category = first(category),
    n_contribs = first(n_contribs),
    n = n(),
    first_commit = min(datetime),
    last_commit = max(datetime),
    lifespan = as.integer(last_commit - first_commit),
    commit_rate = n / lifespan
  ) ->
  oarch_lifespan
```

@@

```{r plot-lifespan-hist}
# TODO: combine with patchwork
# TODO: where does oarch_lifespan come from?
oarch_lifespan %>%
  ggplot(aes(x = lifespan)) +
  geom_histogram(binwidth = 100) +
  scale_x_continuous(breaks = scales::breaks_width(1000)) +
  labs(
    title = "Project lifespan",
    subtitle = "Time between first and latest commit, by repository",
    x = "Days", y = NULL,
    caption = "Data: open-archaeo.info and GitHub API"
  ) +
  theme_oarch()
```

```{r plot-lifespan-vs-rate}
oarch_lifespan %>%
  ggplot(aes(x = lifespan, y = commit_rate, size = n)) +
  geom_point(shape = 21, fill = NA) +
  scale_x_log10() +
  scale_y_log10() +
  labs(
    title = "Project lifespan vs. average commit rate",
    x = "Lifespan (days)", y = "Commits per day", size = "Total commits",
    caption = NULL
  ) +
  theme_oarch()

oarch_lifespan %>%
  ggplot(aes(x = lifespan, y = commit_rate, size = n_contribs)) +
  geom_point(shape = 21, fill = NA) +
  scale_x_log10() +
  scale_y_log10() +
  labs(
    title = NULL,
    x = "Lifespan (days)", y = "Commits per day", size = "Contributors",
    caption = "Data: open-archaeo.info and GitHub API"
  ) +
  theme_oarch()
```

@@

```{r plot-gh_contributors-histogram}
oarch %>%
  drop_na(gh_repo) %>%
  unnest(gh_contribs) %>%
  group_by(gh_repo) %>%
  summarise(n = n(), .groups = "drop") %>%
  ggplot(aes(x = n)) +
  geom_histogram() +
  scale_x_log10() +
  labs(
    title = "Repository contributors",
    x = "Number of contributors", y = "Repositories",
    caption = "Data: open-archaeo.info and GitHub API"
  ) +
  theme_oarch()
```

@@

```{r plot-contrib-dist}
oarch %>%
  unnest(gh_contribs) %>%
  drop_na(contributor) %>%
  group_by(item_name) %>%
  mutate(n_contributors = n()) %>%
  filter(n_contributors > 1) %>%
  mutate(big_boy = ifelse(contributions == max(contributions),
                          "Leading contributor", "Other contributors")) %>% 
  group_by(item_name, big_boy) %>%
  summarise(contributions = sum(contributions)) %>%
  mutate(
    p_contributions = contributions / sum(contributions),
    p_big_boy = first(p_contributions)
  ) %>%
  ungroup() %>%
  mutate(item_name = fct_reorder(item_name, p_big_boy)) %>% 
  ggplot(aes(x = item_name, y = contributions,
             fill = fct_relevel(big_boy, c("Other contributors",
                                           "Leading contributor")))) +
  geom_col(position = "fill", colour = "white", size = 0.1) +
  geom_hline(yintercept = c(0.25, 0.5, 0.75), alpha = 0.5) +
  scale_y_continuous(labels = scales::percent) +
  scale_fill_highcontrast(guide = guide_legend(reverse = TRUE)) +
  labs(title = "Repository contributors by proportion of contributions",
       x = NULL, y = "Contributions",
       fill = NULL,
       caption = "Data: open-archaeo.info and GitHub API") +
  coord_flip() +
  theme_oarch() +
  theme(
    axis.text.y = element_blank(),
    axis.ticks.y = element_blank()
  )
```

@@

```{r plot-issues-hist}
# TODO: combine with comments-hist
oarch %>%
  unnest(gh_issues) %>%
  drop_na(title) %>%
  group_by(repo) %>%
  summarise(n = n(), .groups = "drop") %>%
  ggplot(aes(x = n)) +
  geom_histogram() +
  scale_x_log10() +
  labs(
    title = "Use of issues",
    caption = "Data: open-archaeo.info and GitHub API",
    x = "Number of issues", y = "Repositories"
  ) +
  theme_oarch()
```

```{r plot-comments-hist}
oarch %>%
  unnest(gh_comments) %>%
  drop_na(issue_num) %>%
  group_by(repo) %>%
  summarise(n = n(), .groups = "drop") %>%
  ggplot(aes(x = n)) +
  geom_histogram() +
  scale_x_log10() +
  labs(
    title = "Use of comments",
    caption = "Data: open-archaeo.info and GitHub API",
    x = "Number of comments", y = "Repositories"
  ) +
  theme_oarch()
```

### 2023-06-14 analysis/_07-network.qmd
Merge code chunks into section qmds
Leaving the network analysis unevaluated for now.

{r graph-data}

@@

{r plot-graph-repo-user}

@@

{r plot-graph-repo-repo}
{r graph-user-user}
{r plot-graph-user-user}

@@

{r plot-graph-repo-repo-edge-betweenness}
{r plot-graph-repo-repo-edge-betweenness-dendrogram}

@@

{r summary-graph-user-user}

@@

{r plot-user-betweenness-dist}
{r plot-user-betweenness-activity}

>>>>

```{r graph-data, eval=FALSE}
# TODO: refactor some of this as functions?
oarch |>
  drop_na(gh_repo) |>
  unite(tags, c(tag1, tag2, tag3, tag4, tag5)) |>
  select(gh_repo, category, tags, gh_commits, gh_issues, gh_comments) ->
  oarch_graph

oarch_graph <- bind_rows(
  oarch_graph |>
    unnest(gh_commits) |>
    drop_na(author) |> # Apparently real commits, but we can't connect them
    group_by(repo, author) |>
    summarise(category = first(category), tags = first(tags),
              contrib = "commit", n_contrib = n(), .groups = "drop"),
  oarch_graph |>
    unnest(gh_comments) |>
    drop_na(author) |> # Repos with no comments
    group_by(repo, author) |>
    summarise(category = first(category), tags = first(tags),
              contrib = "comment", n_contrib = n(), .groups = "drop"),
  oarch_graph |>
    unnest(gh_issues) |>
    drop_na(author) |> # Repos with no issues
    group_by(repo, author) |>
    summarise(category = first(category), tags = first(tags),
              contrib = "issue", n_contrib = n(), .groups = "drop")
)

# Cleaning
oarch_graph |>
  rename(user = author) |>
  # Remove non-human users
  filter(!str_detect(user, coll("[bot]"))) |>
  filter(!user %in% c("ghost", "codecov-io", "coveralls", "codecov-commenter",
                      "gitter-badger", "dependabot-support")) |>
  # Remove big, non-archaeology projects that flood the network
  filter(!repo %in% c("Stellarium/stellarium", 
                     "opengisch/QField",
                     "lutraconsulting/input")) ->
  oarch_graph
```

@@

```{r plot-graph-repo-user, eval=FALSE}
oarch_graph_repo_user <- tbl_graph(nodes_repos_users(oarch_graph),
                                   edges_user_to_repo(oarch_graph))

g_repo_user <- oarch_graph_repo_user |>
  ggraph(layout = "graphopt") +
  geom_edge_fan(aes(colour = contrib, alpha = n_contrib)) +
  geom_node_point(aes(colour = node_type)) +
  scale_edge_alpha(trans = "log")

#g_repo_user
girafe(g_repo_user)
```

@@

```{r plot-graph-repo-repo, eval=FALSE}
oarch_graph_repo_repo <- tbl_graph(nodes_repos(oarch_graph, category),
                                   edges_common_users(oarch_graph),
                                   directed = FALSE)

oarch_graph_repo_repo |>
  ggraph(layout = "graphopt") +
  geom_edge_link(aes(alpha = n)) +
  geom_node_point(aes(colour = category)) +
  scale_edge_alpha(trans = "log") +
  labs(edge_alpha = "Common contributors")
```

```{r graph-user-user, eval=FALSE}
oarch_graph_user_user <- tbl_graph(
  nodes = nodes_users(oarch_graph), 
  edges = edges_common_repos(oarch_graph),
  directed = FALSE
)
```
```{r plot-graph-user-user, eval=FALSE}
oarch_graph_user_user |>
  mutate(betweenness = centrality_betweenness(weights = n)) ->
  oarch_graph_user_user

oarch_graph_user_user |> 
  ggraph(layout = "graphopt") +
  geom_edge_link(aes(alpha = n)) +
  # geom_node_point() +
  scale_edge_alpha(trans = "log") +
  labs(edge_alpha = "Common repositories")
```

@@

```{r plot-graph-repo-repo-edge-betweenness, eval=FALSE}
oarch_graph_repo_repo <- tbl_graph(nodes_repos(oarch_graph,category,tags),
                                   edges_common_users(oarch_graph),
                                   directed = FALSE)

oarch_graph_repo_repo |>
  mutate(degree = centrality_degree()) |>
  filter(degree > 1) |>
  mutate(group = as.character(group_edge_betweenness())) ->
  oarch_graph_repo_repo

iplot <- oarch_graph_repo_repo |>
  ggraph(layout = "graphopt") +
  geom_edge_link(aes(alpha = n)) +
  #geom_node_point(aes(colour = group)) +
  #geom_node_text(aes(label = group)) +
  geom_point_interactive(aes(x, y, colour = group, tooltip = repo)) + 
  scale_edge_alpha(trans = "log") +
  labs(edge_alpha = "Common contributors")

girafe(ggobj = iplot)
```

```{r plot-graph-repo-repo-edge-betweenness-dendrogram, eval=FALSE}
oarch_graph_repo_repo <- tbl_graph(nodes_repos(oarch_graph),
                                   edges_common_users(oarch_graph),
                                   directed = TRUE)

oarch_graph_repo_repo |>
  mutate(degree = centrality_degree()) |>
  filter(degree > 1) |>
  mutate(group = as.character(group_edge_betweenness())) ->
  oarch_graph_repo_repo

xx <- igraph::edge.betweenness.community(oarch_graph_repo_repo)
xx <- as.dendrogram(xx)
plot(xx)

ggraph(xx, 'dendrogram', height = height)
```

@@

```{r summary-graph-user-user, eval=FALSE}
oarch_graph_user_user |>
  mutate(degree = centrality_degree(), 
         betweenness = centrality_betweenness(weights = n), 
         strength = centrality_degree(weights = n)) |>
  filter(degree > 0) |> 
  data.frame()
```

@@

```{r plot-user-betweenness-dist, eval=FALSE}
oarch_graph_user_user |>
  mutate(
    betweenness = centrality_betweenness(weights = n)
  ) |>
  data.frame() |>
  ggplot(aes(x = betweenness)) +
  geom_density(bw = "nrd") +
  scale_x_log10(labels = scales::number)
```

```{r plot-user-betweenness-activity, eval=FALSE}
oarch_graph_user_user |>
  mutate(
    degree = centrality_degree(), 
    betweenness = centrality_betweenness(weights = n), 
    strength = centrality_degree(weights = n)
  ) |>
  as_tibble() |> 
  left_join(oarch_contributors, by = c("user" = "contributor")) |>
  ggplot(aes(betweenness, contribs)) +
  geom_point() +
  scale_x_log10() +
  scale_y_log10() +
  geom_smooth(method = "lm")
```
